Crystal as a thermodynamic computer

### 2024-09-04 paper/main.tex
update

\title{Surface programmable materials}

@@

The computational process of TCs utilizes the tendency of a physical system to evolve towards the state with the lowest free energy.
By manipulating the minimum free energy state of the system, the computation can be performed.
The change of free energy minimum state can be achieved by changing the distribution of the particles in the system, such as in the case of DNA machines~\cite{Feynman2018} or lowering the temperature of the system, such as in the case of Ising machines~\cite{boixo2013experimental}.
The former is known as the Brownian machine, which evolves the system toward the direction of the entropy increase, while the latter are energy based, which evolves the system toward the direction of the energy decrease.
\jinguo{What category are the memristors?}

@@

Intrigued by the question of how to thermalize and lower the energy of an Ising machine, we propose a new type of TC that has a crystalline structure.
The crystalline structure is based on a universal elementary cellular automata, and is translational invariant in all spacial dimensions.
The translational invariance not only makes the crystal easier to be constructed in nature, but also enables the study of the scaling behavior the thermalization process.
In the rest of this paper, we first show that the crystalline structure can be programmed on its surfaces to perform universal computation \Cref{sec:Overview-of-main-results}.
Then we use numerical results in \Cref{sec:numerical-result} to show that if we thermalize the crystal from the so-called ``deterministic direction'', the crystal can thermalize in time almost linear to the size of the crystal, i.e. efficient computation can be performed.

@@

\section{Overview of the computation model}\label{sec:Overview-of-main-results}
In this article, We develop a new type of TC shown in \Cref{Surface-programmable-thermodynamic computer}(a), which features a crystalline structure based on a universal elementary cellular automata, makes it easier to be constructed in real life.

@@

    % Text Node
    \draw (379,111.8) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \cdots $};
    % Text Node
    \draw (379,141.8) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \cdots $};
    % Text Node
    \draw (379,172.09) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \cdots $};
    % Text Node
    \draw (379,241.8) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \cdots $};
    % Text Node
    \draw (283,193.97) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \vdots $};
    % Text Node
    \draw (322.5,193.97) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \vdots $};
    % Text Node
    \draw (363,193.97) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \vdots $};
    % Text Node
    \draw (420.1,194) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \vdots $};
    % Text Node
    \draw (382.4,194.52) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \ddots $};
    % Text Node
    \draw (292,45) node [anchor=north west][inner sep=0.75pt]   [align=left] {${Heat\ Zone}$};
    % Text Node
    \draw (220.87,45.67) node [anchor=north west][inner sep=3pt]   [align=left] {\large {$\displaystyle \mathtt{(a)}$}};
    % Text Node
    \draw (490.67,121.47) node [anchor=north west][inner sep=3pt]   [align=left] {\large {$\displaystyle \mathtt{(b)}$}};
    % Text Node
    \draw (492.43,150.41) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny {\fontfamily{ptm}\selectfont p}}};
    % Text Node
    \draw (479.75,168.95) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\fontfamily{ptm}\selectfont {\tiny q}}};
    % Text Node
    \draw (492.83,188.12) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny {\fontfamily{ptm}\selectfont r}}};
    % Text Node
    \draw (525.34,167.95) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny {\fontfamily{ptm}\selectfont q'}}};
    % Text Node
    \draw (504.00,168.45) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny {\fontfamily{ptm}\selectfont A}}};

@@

    \end{tikzpicture}
    \caption{Surface programmable thermodynamic computer (TC) and rule110 automata gadget. $\mathtt{(a)}$ Example of a surface programmable TC with a moving heat bath slowly doing ``calculation''. White circles are spins and black lines represent interaction between spins, while red rectangles represent cellular automata gadget. $\mathtt{(b)}$ Example of a rule110 gadget. Here spins $p$, $q$, and $r$ represent the left, middle, and right states in the cellular automata, respectively, while $q^\prime$ represents the result computed by the cellular automata, and $A$ is an auxiliary spin. We define a spin pointing up as representing logical 1 and pointing down as representing logical 0. \jinguo{why capital A? q' not clear.}}
    \label{Surface-programmable-thermodynamic computer}
\end{figure}

@@

The input layer consists of the leftmost spin, while the output layer consists of the rightmost spin. The entire computing process can be described by gradually moving the external heat bath, which creates a temperature gradient and, consequently, a heat zone, from the input layer to the output layer. By utilizing thermal fluctuations from the heat bath, the spins in different layers sequentially settle into low-energy states within the state space, effectively completing the iterative computation of the cellular automata.

@@

\subsection{Elementary cellular automata}\label{elementary-cellular-automata}
In cellular automata (CA), `space' is a regular lattice of `cells', each of which contains one of a small allowed set of integers.
Only cells that are close together interact in one `time-step'-the time evolution is given by a rule that looks at the contents of a few neighbouring cells, and decides what should change. At each step, this local rule is applied everywhere simultaneously~\cite{}.
Depending on different rules, some CAs can be more complex than others.
The most interesting category is those proved to be universal~\cite{Berlekamp2004}, meaning that they can simulate any computer program. The best-known examples in two-dimensional include Conway's~\cite{Games1970} ``Game of Life'', Margolus's reversible cellular automata~\cite{Margolus1984} and its quantum version~\cite{Arrighi2012}.
The simplest universal CA is a 1-dimensional cellular automata, namely, the rule 110 CA~\cite{Cook2009}.
This CA is an elementary CA that defined on a one dimensional grid, each cell has two possible states (labeled by 0 and 1). The rule of an elementary CA to determine the state of the cell in next generation depends only on the current state of the cell and its two immediate neighbors. The number of possible elementary CA rules is the same as the number of 3-input Boolean functions, which is $2^{2^3} = 256$. The Boolean function for rule 110 is shown in the following table:

@@

The name of the rule comes from the binary representation of the rule number, which is 110 in decimal.
Let us denote the states of the three cells as $p$, $q$, and $r$, from left to right, respectively. The rule 110 can be expressed as the following Boolean function:
\begin{equation}\label{eq:110rule}
    q^\prime = r \land \neg (q \land \neg p)
\end{equation}
where $q^\prime$ is the state of the center cell in the next generation.

@@

Previous studies have shown that the Ising model is universal for computation~\cite{Aadit2022, onizawa2020design, camsari2017stochastic}, meaning that any logic circuit can be encoded into the Ising model and solved by finding the ground state of the Ising model. A universal set logic gate gadgets are designed, by concatenating which any logic circuit can be constructed.
For example the logic expression $z = \neg (x \land y)$ can be encoded into the Ising model by the following Hamiltonian:
\begin{equation}
    \begin{split}
        H_{c = a \land b} &= s_a s_b -2 s_a s_c -2 s_bs_c + s_a + s_b -2 s_c\\
        H_{z = \neg c} &= s_c s_z\\
        H_{z = \neg (a \land b)} &= H_{z = \neg c} + H_{c = a \land b}
    \end{split}
\end{equation}
where, the variable $c$ is an ancilla spin.
A cellular automata can be viewed as a translational invariant logic circuit, and thus can be encoded into the Ising model. The ground states of the Ising model encode the correspondence between the inputs and the outputs of the cellular automata.
Directly synthesizing the logic circuit for the universal $110$ rule (\Cref{eq:110rule}) requires at least 3 ancilla spins, which is far from optimal. By converting the gadget finding problem to a linear programming problem~\cite{}, we propose a provable optimal method to encode any cellular automata rule into the Ising model, which is described in \Cref{sec:gadget-design}.
The proposed gadget for the $110$ rule is shown in \Cref{Surface-programmable-thermodynamic computer}(b), which only requires 1 ancilla spin.
The interaction energy and onsite energy of the rule 110 gadget are given below, where $p, q, r$ and inputs and $q'$ is the output in \Cref{Surface-programmable-thermodynamic computer}(b), respectively, while spin $A$ is the auxiliary spin.
\begin{equation}
J = \begin{blockarray}{cccccc}
& p & q & r & q^\prime & A\\
\begin{block}{c(ccccc)}
p & ~\cdot~ & ~1~ & ~1~ & ~2~ & ~3~\\
q & \cdot & \cdot & 2 & 2 & 5\\
r & \cdot & \cdot & \cdot & 2 & 5\\
q^\prime & \cdot & \cdot & \cdot & \cdot & 6\\
A & \cdot & \cdot & \cdot & \cdot & \cdot\\
\end{block}
\end{blockarray}
~~, ~~h = \begin{blockarray}{cc}
& \phantom{x} \\
\begin{block}{c(c)}
p & ~1~\\
q & 2\\
r & 2\\
q^\prime & 2\\
A & 5\\
\end{block}
\end{blockarray}
\end{equation}
Since a copy gate\jinguo{do we really use copy gates?} can be simply described as two spins with an interaction $J_{u, v}=1$, and by applying the method of connecting logic gates~\cite{Aadit2022}, we can concatenate the rule 110 gadget with some copy gates, ultimately forming the TC shown in [Fig. \ref{Surface-programmable-thermodynamic computer}(a)]. The gadget we constructed based on the Rule 110 cellular automata
naturally possesses Turing completeness. Therefore, this 2-dimensional TC is capable of universal computing.

@@

The computational process of the cellular automata is divided into three steps: determining the state of each input cell, iteratively applying the transition rules to generate the next generation of cells, and identifying a certain generation of cells as the output. In fact, the width of the time axis in the Ising TC corresponds to the number of iterations. Hence, the smallest layer of spins on the time axis should be the input layer, and the largest layer of spins should be the output layer.

Given that each ground state of the TC corresponds to a valid computational process of the cellular automata for a given input, the computational process of the TC is essentially reduced to finding the ground state of the TC with the spin configuration in the input layers fixed.

@@

\ym{This part should be rewrote in future..My words are so poor.}

@@

In practice, temperature changes cannot occur instantaneously, and the temperature distribution within the heating zone may not be uniform. Thus, it is necessary to account for the effective width of the Heat Zone when the the heat bath are not ideal.

Suppose the middle of the heat bath is at $middle(t)$, and the temperature at the region with a Time coordinate $x$ can be represent as $Temp(x) \propto \Lambda^{-|x - middle(t)|}$. Then

\begin{equation}
    e^{-\Delta E \Lambda^{cW}} \sim \epsilon \sim \frac{W}{T}
\end{equation}

Where $c$ is the empirical constant for effective width and can be determined during the experiment. The total annealing time should be

@@

However, we fail to derive an explicitly expression for the time complexity in this direction. 
Numerical result shows that computing complexity is linear to $n$. Thus a total time
for the computation should scale as $nm\log^2(m)e^{(1-\lambda)^{-1}}$

\subsubsection{Testing model through equilibrilium SA}\label{temperature-gradient-and-energy-gradient}

We used simulated annealing to mimic the heating-annealing process.
However, classical simulated annealing can only handle the situation
where the temperature is space-invariant. Thus, a scheme change is needed.

One can simply
make the following modification to the transition probability in the
markov process and return to the situation, called energy-gradient scheme, where the temperature is space-invariant.

\begin{equation}
e^{-\frac{\Delta E_k}{T\lambda^{ct+k}}} = e^{-\frac{\Delta E_k(\frac{1}{\lambda})^k}{T\lambda^{ct}}} = e^{-\frac{\Delta E_k \Lambda^k}{T(t)}}
\end{equation}

Where we denote $\Lambda = \frac{1}{\lambda}$ and the time complexity in $\Lambda$ form is apparently
$T_{total} \sim \frac{m}{W}\log^2(\frac{m}{W})e^{(\Lambda-1)^{-1}}$

\subsection{Algorithm beyond temperature
gradient}\label{algorithm-beyond-temperature-gradient}

Applying a huge energy gradient or temperature gradient to this material
seems not a good idea when the system become larger.

Inspired by the way clothes are ironed, we conceived an idea called wave-packet scheme, that slowly
scanning a wave-shaped temperature curve from one end to the other. This
method is widely used in studying the heating-recrystallization
properties of real materials~\cite{Zhang2014}.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{../notes/images/zhang2014.png}
\caption{Method to scan the wave packet}
\end{figure}

There are various specific choices for the wave packet, such as
power-law, exponential, and others. In this context, we primarily focus
on a exponential wave packet, which can be directly linked to the
concept of energy gradient.

A expoential wave packet with central position $middle(t)$, amplitude
$A$ and base $\Lambda$, can be express as following.

\[
T(k,t) = A\Lambda^{|k-middle(t)|}
\]

So slowly moving the wave packet means slowly changing $middle(t)$

Let's review the concept \textbf{active zone} defined earlier. When the
temperature of the system is defined as
$T_{tg}(k, t) = T\lambda^{ct-k}$, atoms in the active zone are those
whose layer $k$ belongs to $k<ct - \frac{W}{2}$, where $W$ are
given by
$e^{-\frac{\Delta E_{max}}{\lambda^{\frac{W}{2}}}}\leq \epsilon$. For
those $k<ct - \frac{W}{2}$, the probability that SA would flip an
atom's state is no more than
$e^{-\frac{\Delta E_{max}}{T\lambda^{ct}\lambda^{-ct + \frac{W}{2}}}}=e^{-\frac{\Delta E_{max}}{T\lambda^{\frac{W}{2}}}}\leq \epsilon^{\frac{1}{T}}$,
which means these previous layer would stay ``frozen''. For those layer
much deeper than $k = ct$, SA would just randomly flip them. These
layer won't affect layers in active zone.

Now turn our sight into the wave packet model, we can construct a
similar concept \textbf{heat zone}. Layers in the heat zone are those
$A\lambda^{|k-middle(t)|}\ge \epsilon\rightarrow middle(t) + \frac{\log(\epsilon)}{\log(\Lambda)}< k < middle(t) - \frac{\log(\epsilon)}{\log(\Lambda)}$.

The layers that shallower than
$middle(t) + \frac{\log(\epsilon)}{\log(\Lambda)}$ correspond to
layers shallower than $ct-\frac{W}{2}$ in temperature gradient model,
since they are both stay ``frozen''; layers between
$middle(t) +\frac{\log(\epsilon)}{\log(\Lambda)}$ and $middle(t)$
are effectively cooled down, which correspond to the \textbf{active
zone}; layers deeper than $middle(t)$ are either randomly fliped or
stay ``frozen'', but we don't care their state.

Therefore, the wave packet model are just the same as the temperature
gradient model, and is more likely to be realized physically.

\subsubsection{Some details about non-equilibrium simulated
annealing}\label{some-details-about-non-equilibrium-simulated-annealing}

The classical simulated annealing can only deal with the situation that
the temperature is space-invariant. Although there are many ways to
simulated a real-world system with temperature gradient, such as
Molecular Dynamics \cite{Bai2015, Deng2006}, but
these methods mainly depend on dynamically adjusting the atomic
velocities to achieve a temperature gradient, which can't be directly
used in our model.

@@

When we flip one atom's state, there would be some energy difference
$\Delta E_{i}$ in field $i$ with temperature $T_i$. The
probability to accept this flip would be

@@

It is easily to observe that when the system returns to
heat-equilibrium, the probability reverts to the classical HeatBath
acceptance probability.
\section{Numerical simulation}\label{sec:numerical-result}

@@

Given that simulating a cellular automata with a deep computational layer on the Ising model requires a large number of spins, making large-scale numerical simulations impractical, we opted to introduce a similar but simpler energy model to observe the behavior on a larger scale.

Let us define a model Hamiltonian on a 2D lattice of size $S \times T$ as
\begin{equation}\label{eq:toy-hamiltonian}
H = \sum_{t=1}^{T-1}\sum_{i=1}^{S} (n_{i+1, t} \land \neg(n_{i, t} \land \neg n_{i-1, t})) \veebar n_{i,t+1},
\end{equation}
where $n_{i, t} \in \{0, 1\}$ is the state of the cell at row $i$ and column $t$, the boundary is periodic in the space direction, i.e. $n_{0, t} = n_{S, t}$.
Its ground state with energy $0$ encodes a $110$ cellular automata, where the direction of computation is from the left to right.
\subsection{Numerical result}

@@

@@

We can encode the Rule 110 cellular automata into a Weighted Maximum
Independent Set Problem, with blue vertices assigned a weight of 1 and
red vertices assigned a weight of 2, as follows.

@@

The states of vertex \textbf{1}, vertex \textbf{3}, and vertex
\textbf{8} represent the states of the \textbf{middle}, \textbf{left},
and \textbf{right} cells of the automata's \textbf{input},
respectively. If the input value of a cell is 1, then the corresponding
vertex must be in the MWIS solution; otherwise, it is not. Vertex
\textbf{12} corresponds to the automata's \textbf{output}. If the
automata output is 1, then vertex 12 is in the MWIS solution;
otherwise, it is not.

In the automata diagram, the above gadget is equivalent to:

@@

There are exactly \textbf{8} different MWIS solutions in this graph (the
weighted size of each MWIS solution is 7), each corresponding to one of
the \textbf{8} possible outputs of the automata. We list them as
follows.
\includegraphics[width=\columnwidth]{../notes/images/gadget110.png}

@@

The above gadget depicts a two-layer cellular automata. The vertices in
blue, red, green and black have weights of 1, 2, 3 and 4, respectively.
In the automata diagram, the above gadget is equivalent to:

>>>>

\title{Crystal as a thermodynamic computer}

@@

The computational process of TCs utilizes the tendency of a physical system to evolve towards the state with the lowest free energy.
By manipulating the minimum free energy state of the system, the computation can be performed.
The change of free energy minimum state can be achieved by changing the distribution of the particles in the system, such as in the case of DNA machines~\cite{Feynman2018} or lowering the temperature of the system, such as in the case of Ising machines~\cite{boixo2013experimental}.
The former is known as the Brownian machine, which evolves the system toward the direction of the entropy increase, while the latter are energy based, which evolves the system toward the direction of the energy decrease.
\jinguo{What category are the memristors?} \ym{Memristors are energy based, which evolves the system toward the direction of the energy decrease.}

@@

Intrigued by the question of how to thermalize and lower the energy of an Ising machine, we propose a new type of TC that has a crystalline structure.
The crystalline structure is based on a universal elementary cellular automaton, and is translational invariant in all spacial dimensions.
The translational invariance not only makes the crystal easier to be constructed in nature, but also enables the study of the scaling behavior the thermalization process.
In the rest of this paper, we first show that the crystalline structure can be programmed on its surfaces to perform universal computation \Cref{sec:Overview-of-main-results}.
Then we use numerical results in \Cref{sec:numerical-result} to show that if we thermalize the crystal from the so-called ``deterministic direction'', the crystal can thermalize in time almost linear to the size of the crystal, i.e. efficient computation can be performed.

@@

\section{Overview of the computation model}\label{sec:Overview-of-main-results}
In this article, We develop a new type of TC shown in \Cref{Surface-programmable-thermodynamic computer}(a), which features a crystalline structure based on a universal elementary cellular automaton, makes it easier to be constructed in real life.

@@

    % Text Node
    \draw (379,111.8) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \cdots $};
    % Text Node
    \draw (379,141.8) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \cdots $};
    % Text Node
    \draw (379,172.09) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \cdots $};
    % Text Node
    \draw (379,241.8) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \cdots $};
    % Text Node
    \draw (283,193.97) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \vdots $};
    % Text Node
    \draw (322.5,193.97) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \vdots $};
    % Text Node
    \draw (363,193.97) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \vdots $};
    % Text Node
    \draw (420.1,194) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \vdots $};
    % Text Node
    \draw (382.4,194.52) node [anchor=north west][inner sep=0.75pt]   [align=left] {$\displaystyle \ddots $};
    % Text Node
    \draw (292,45) node [anchor=north west][inner sep=0.75pt]   [align=left] {${Heat\ Zone}$};
    % Text Node
    \draw (220.87,45.67) node [anchor=north west][inner sep=3pt]   [align=left] {\large {$\displaystyle \mathtt{(a)}$}};
    % Text Node
    \draw (490.67,121.47) node [anchor=north west][inner sep=3pt]   [align=left] {\large {$\displaystyle \mathtt{(b)}$}};
    % Text Node
    \draw (492.00,150.41) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny {$p$}}};
    % Text Node
    \draw (479.75,168.95) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny {$q$}}};
    % Text Node
    \draw (492.45,188.12) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny {$r$}}};
    % Text Node
    \draw (525.00,166.95) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny {$q'$}}};
    % Text Node
    \draw (504.00,168.45) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny {$A$}}};

@@

    \end{tikzpicture}
    \caption{Surface programmable thermodynamic computer (TC) and rule110 automaton gadget. $\mathtt{(a)}$ Example of a surface programmable TC with a moving heat bath slowly doing ``calculation''. White circles are spins and black lines represent interaction between spins, while red rectangles represent cellular automaton gadget. $\mathtt{(b)}$ Example of a rule110 gadget. Here spins $p$, $q$, and $r$ represent the left, middle, and right states in the cellular automaton, respectively, while $q’$ represents the result computed by the cellular automaton, and $A$ is an auxiliary spin. We define a spin pointing up as representing logical 1 and pointing down as representing logical 0. \jinguo{why capital A? q' not clear.}}
    \label{Surface-programmable-thermodynamic computer}
\end{figure}

@@

The input layer consists of the leftmost spin, while the output layer consists of the rightmost spin. The entire computing process can be described by gradually moving the external heat bath, which creates a temperature gradient and, consequently, a heat zone, from the input layer to the output layer. By utilizing thermal fluctuations from the heat bath, the spins in different layers sequentially settle into low-energy states within the state space, effectively completing the iterative computation of the cellular automaton.

@@

\subsection{Elementary cellular automaton}\label{elementary-cellular-automaton}

\jinguo{general CA, then the 1D CA.}
\ym{I haven't made change in this section}

An elementary cellular automaton is a 1-dimensional cellular
automaton, where there are two possible states (labeled by 0 and 1). The rule to
determine the state of the cell in next generation depends only on the
current state of the cell and its two immediate neighbors.

There are $8 = 2^3$ possible configurations for a cell and its two
immediate neighbors. Different elementary cellular automaton are only
different from their translation rules. There are only $2^8 = 256$
different rules, so do the automatons.

If we put each possible current configurations in order: 111, 110,
\ldots, 001, 000, and put the resulting state under them. We then get an
integer in its binary representations. Then this integer is taken to be
the rule number of the automaton. For example, rule 110.

Given that $110_d = 01101110_2$, so rule 110 is defined by the
translation rule:

@@

In boolean form, it can be expressed as:
\begin{equation}\label{eq:110rule}
    q^\prime = r \land \neg (q \land \neg p)
\end{equation}

Rule 110 has been shown to be Turing Complete~\cite{Cook2009}, and thus capable of universal
computation.

@@

Previous studies have shown that the Ising model is universal for computation~\cite{Aadit2022, onizawa2020design, camsari2017stochastic}, meaning that any logic circuit can be encoded into the Ising model and solved by finding the ground state of the Ising model. A universal set logic gate gadgets are designed, by concatenating which any logic circuit can be constructed.
For example the logic expression $z = \neg (x \land y)$ can be encoded into the Ising model by the following Hamiltonian:
\begin{equation}
    \begin{split}
        H_{c = a \land b} &= s_a s_b -2 s_a s_c -2 s_bs_c + s_a + s_b -2 s_c\\
        H_{z = \neg c} &= s_c s_z\\
        H_{z = \neg (a \land b)} &= H_{z = \neg c} + H_{c = a \land b}
    \end{split}
\end{equation}
where, the variable $c$ is an ancilla spin.
A cellular automaton can be viewed as a translational invariant logic circuit, and thus can be encoded into the Ising model. The ground states of the Ising model encode the correspondence between the inputs and the outputs of the cellular automaton.
Directly synthesizing the logic circuit for the universal $110$ rule (\Cref{eq:110rule}) requires at least 3 ancilla spins, which is far from optimal. By converting the gadget finding problem to a linear programming problem~\cite{}, we propose a provable optimal method to encode any cellular automaton rule into the Ising model, which is described in \Cref{sec:gadget-design}.
The proposed gadget for the $110$ rule is shown in \Cref{Surface-programmable-thermodynamic computer}(b), which only requires 1 ancilla spin.
The interaction energy and onsite energy of the rule 110 gadget are given below, where $p, q, r$ and inputs and $q'$ is the output in \Cref{Surface-programmable-thermodynamic computer}(b), respectively, while spin $A$ is the auxiliary spin.
\begin{equation}
J = \begin{blockarray}{cccccc}
& p & q & r & q^\prime & A\\
\begin{block}{c(ccccc)}
p & ~\cdot~ & ~1~ & ~1~ & ~2~ & ~3~\\
q & \cdot & \cdot & 2 & 2 & 5\\
r & \cdot & \cdot & \cdot & 2 & 5\\
q^\prime & \cdot & \cdot & \cdot & \cdot & 6\\
A & \cdot & \cdot & \cdot & \cdot & \cdot\\
\end{block}
\end{blockarray}
~~, ~~h = \begin{blockarray}{cc}
& \phantom{x} \\
\begin{block}{c(c)}
p & ~1~\\
q & 2\\
r & 2\\
q^\prime & 2\\
A & 5\\
\end{block}
\end{blockarray}
\end{equation}
Since a copy gate\jinguo{do we really use copy gates?} \ym{No need in practice, but indeed used in Fig.1a} can be simply described as two spins with an interaction $J_{u, v}=1$, and by applying the method of connecting logic gates~\cite{Aadit2022}, we can concatenate the rule 110 gadget with some copy gates, ultimately forming the TC shown in [Fig. \ref{Surface-programmable-thermodynamic computer}(a)]. The gadget we constructed based on the Rule 110 cellular automaton
naturally possesses Turing completeness. Therefore, this 2-dimensional TC is capable of universal computing.

\subsection{Simulated Annealing}

It is known that when the system and heat bath are at thermal equilibrium, 
the probability of a spin configuration $|\mathbf{s}\rangle$ is given by the Boltzmann distribution:

\begin{equation}
    |\mathbf{s}\rangle = \frac{1}{Z}\exp\left(- \frac{E_{|\mathbf{s}\rangle}}{k_bT}\right)
\end{equation}

where $Z = \sum_{|\mathbf{s}\rangle} \exp\left(-\frac{E_{|\mathbf{s}\rangle}}{k_bT}\right)$ is the partition function, 
$E_{|\mathbf{s}\rangle}$ is the energy of the each of $2^n$ spin configuration, labeled by $|\mathbf{s}\rangle = |s_1, s_2, \cdots, s_n\rangle$, $k_b$ is the Boltzmann constant, and $T$ is the temperature of the heat bath. 
In the following, we set $k_b = 1$.

We employ the Metropolis-Hastings algorithm~\cite{metropolis1953equation} to characterize the thermodynamic process of the TC system as modulated by the heat bath.
This algorithm perform a random walk in the state space of the TC, each step we try to flip a spin,
and the probability of accepting a move from state $|\mathbf{s}\rangle$ to $|\mathbf{s}^\prime\rangle$ is given by:

\begin{equation}
    P_{\text{accept}} = \frac{1}{1+\exp\left(\frac{E_{|\mathbf{s}^\prime\rangle} - E_{|\mathbf{s}\rangle}}{T}\right)}
\end{equation}

This accepting probability follows from the glauber dynamics~\cite{glauber1963time}.
It can be shown that this process is markovian and ergodic, and the system will eventually reach the thermal equilibrium.
To determine the speed of the convergence, we should notice the spectral gap of the transition matrix of the Markov chain, which is defined as

\begin{equation}
    \Delta = 1 - \frac{\lambda_2}{\lambda_1}
\end{equation}

where $\lambda_1$ and $\lambda_2$ are the largest and second largest eigenvalues of the transition matrix, respectively.
In order to sample from the distribution $\epsilon$ close to the equilibrilium distribution, the number of steps required is $O(\log_{1-\Delta} \epsilon)$.

@@

The computational process of the cellular automaton is divided into three steps: determining the state of each input cell, iteratively applying the transition rules to generate the next generation of cells, and identifying a certain generation of cells as the output. In fact, the width of the time axis in the Ising TC corresponds to the number of iterations. Hence, the smallest layer of spins on the time axis should be the input layer, and the largest layer of spins should be the output layer.

Given that each ground state of the TC corresponds to a valid computational process of the cellular automaton for a given input, the computational process of the TC is essentially reduced to finding the ground state of the TC with the spin configuration in the input layers fixed.

@@

@@

In practice, temperature changes cannot occur instantaneously, and the temperature distribution within the heating zone may not be uniform.
Thus, we come up with a more realistic model for the block annealing process.

Now suppose the heat bath only connects to a small rectangle region of the system,
covers the region encompassing all time coordinates between $[middle(t)-\frac{\Delta_L}{2}, middle(t)+\frac{\Delta_L}{2}]$, where $\Delta_L$ is a small constant.
The heat radiation is conducted to other regions, thereby inducing a temperature gradient,
which can be modeled as $T(x) \propto \Lambda^{-|x - middle(t)|}$, where $T(x)$ is the temperature at the region with a time coordinate $x$, and $\Lambda$ is a constant.

The block annealing process then can be reconstruct as slowly moving
the middle of the heat bath from one end to the other. Defining $\Delta E_{max}$ to be the
largest possible energy change when a spin is fliped.
The effective heat zone should then be $[y-W, y]$, where position $y$ satisfying 
$\exp(\frac{\Delta E_{max}}{T(y)}) = 0.1$ and $y<middle(t)$. Since the layers
in fronts of the effective heat zone can only be fliped with a $\epsilon$ probability if the local energy would increase. 
The effective width $W$ then should scale as $\exp(-\Delta E_{max} \Lambda^{cW}) \sim \epsilon \sim \frac{W}{T}$, where $c$ is the empirical constant for effective width and can be determined during the experiment.

The total annealing time can be calculated as

@@

However, we fail to derive an explicitly expression for the time complexity in this direction. 
Numerical result shows that computing complexity is linear to $n$. Thus a total time
for the computation should scale as $nm\log^{1 + \frac{1}{c\ln(\Lambda)}}(m)$

\subsubsection{Testing model with non-equilibrium simulated
annealing}\label{some-details-about-non-equilibrium-simulated-annealing}

The classical simulated annealing can only deal with the situation that
the temperature is space-invariant. However, we need to borden this concept
due to the temperature gradient. Although there are many ways to
simulated a real-world system with temperature gradient, such as
Molecular Dynamics \cite{Bai2015, Deng2006}, but
these methods mainly depend on dynamically adjusting the atomic
velocities to achieve a temperature gradient, which can't be directly
used in our model.

@@

When we flip one spin, there would be some energy difference
$\Delta E_{i}$ in field $i$ with temperature $T_i$. The
probability to accept this flip would be

@@

It is easily to observe that when the system returns to
heat-equilibrium, the probability reverts to the classical HeatBath
acceptance probability. We also provided a more general proof by constructing another equivalent model in Appendix \ref{proof-of_non-equilibrium-SA}.

\section{Numerical simulation}\label{sec:numerical-result}

@@

Given that simulating a cellular automaton with a deep computational layer on the Ising model requires a large number of spins, making large-scale numerical simulations impractical, we opted to introduce a similar but simpler energy model to observe the behavior on a larger scale.

Let us define a model Hamiltonian on a 2D lattice of size $S \times T$ as
\begin{equation}\label{eq:toy-hamiltonian}
H = \sum_{t=1}^{T-1}\sum_{i=1}^{S} (n_{i+1, t} \land \neg(n_{i, t} \land \neg n_{i-1, t})) \veebar n_{i,t+1},
\end{equation}
where $n_{i, t} \in \{0, 1\}$ is the state of the cell at row $i$ and column $t$, the boundary is periodic in the space direction, i.e. $n_{0, t} = n_{S, t}$.
Its ground state with energy $0$ encodes a $110$ cellular automaton, where the direction of computation is from the left to right.
\subsection{Numerical result}

@@

\section{Proof of the applicability of non-equilibrium SA}\label{proof-of_non-equilibrium-SA}

Assume there is another model where the temperature at the region with 
time coordinate $x$ at step $t$ is $T(x, t) = T_0\Lambda^{x - \eta t}$, where 
$\eta$ is cooling constant and $T_0$ satisfies $\exp(\frac{\Delta E_{max}}{T_0})\approx 1$, 
which means all the system stays in a relative high temperature initially. 
We called this model as temperature gradient model.

The cooling process starts at $t=0$ and ends at $t=t_{end}$, where $t_{end}$ satisfies
$\exp(\frac{\Delta E_{max}}{T_0\Lambda^{T - \eta t_{end}}}) \gg 1$, representing there wouldn't be any heat
fluctuations in the system. By noticing that we can define the effective heat zone
similarly to the moving heat bath model and then evaluating the time complexity, one can show that these two model are
equivalent.

The temperature gradient model can then be easily transform to
the energy gradient model in which equilibrilium SA can be applied by rescale local interacting and onsite energy.
The markov process can be treated with uniform temperature while rescale the energy associated with spin with Time-coordinate $x$
to be $E_{energy-gradient}(x) = E_{temp-gradient}(x)\Lambda^{-x}$. The acceptance probability is
invariant under this transformation.

\begin{equation}
    \exp\left({\frac{\Delta E_x}{T\Lambda^{x-\eta t}}}\right) = \exp\left({\frac{\Delta E_x\Lambda^{-x}}{T\Lambda^{-\eta t}}}\right) = \exp\left({\frac{\Delta E_x \Lambda^{-x}}{T(t)}}\right)
\end{equation}

@@

We can encode the Rule 110 cellular automaton into a Weighted Maximum
Independent Set Problem, with blue vertices assigned a weight of 1 and
red vertices assigned a weight of 2, as follows.

@@

The states of vertex \textbf{1}, vertex \textbf{3}, and vertex
\textbf{8} represent the states of the \textbf{middle}, \textbf{left},
and \textbf{right} cells of the automaton's \textbf{input},
respectively. If the input value of a cell is 1, then the corresponding
vertex must be in the MWIS solution; otherwise, it is not. Vertex
\textbf{12} corresponds to the automaton's \textbf{output}. If the
automaton output is 1, then vertex 12 is in the MWIS solution;
otherwise, it is not.

In the automaton diagram, the above gadget is equivalent to:

@@

There are exactly \textbf{8} different MWIS solutions in this graph (the
weighted size of each MWIS solution is 7), each corresponding to one of
the \textbf{8} possible outputs of the automaton. We list them as
follows.
\includegraphics[width=\columnwidth]{../notes/images/gadget110.png}

@@

The above gadget depicts a two-layer cellular automaton. The vertices in
blue, red, green and black have weights of 1, 2, 3 and 4, respectively.
In the automaton diagram, the above gadget is equivalent to:
